### 2020-06-14 manuscript/PLOS-MSc-Thesis.tex
fix Jay's affiliation and reformat the Abstract

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Gene regulatory networks producing a striped band of gene 
expression} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Juan Camilo Arboleda-Rivera\textsuperscript{1\Yinyang},
Gloria Machado-Rodríguez\textsuperscript{1\Yinyang},
Boris Rodríguez-Rey\textsuperscript{2\textcurrency},
Jayson Gutiérrez\textsuperscript{3},
% Name5 Surname\textsuperscript{2\ddag},
% Name6 Surname\textsuperscript{2\ddag},
% Name7 Surname\textsuperscript{1,2,3*},
% with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Instituto de Biología, Universidad de Antioquia, Medellín, Antioquia, Colombia
\\
\textbf{2} Instituto de Física, Universidad de Antioquia, Medellín, Antioquia, Colombia
\\
\textbf{3} Affiliation Dept/Program/Center, Ghent University, Ghent, State, Belgium
% \textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta 
erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante 
cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend 
volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar 
non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc
convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim 
ultrices, ultrices elit pulvinar, volutpat risus.

@@

Our understanding of how GRNs process information has increased thanks to the
concept and theory of network motifs, which are defined as patterns of
interconnections occurring in networks at numbers that are significantly higher
than those in randomized networks \cite{Milo2002}. This novel approach, along
with mathematical modeling and computational methods have been shown to be a
powerful tool to study GRNs and the predictions of studies using these methods
have been validated by experimental studies
\cite{Kalir2005,Kalir2004,Mangan2003a,ODonnell2005,Basu2005}.

@@

The classification of GRNs showed that in our initial set of 2061 GRNs exists a 
number of 714 different network topologies, each one of them containing a 
different number of GRNs (Fig~\ref{fig:distopol}). This abundances distribution 
was not produced by chance (Shannon entropy = 8.52963, 95\% confidence interval 
= [10.8915, 10.9275]), and instead it shows a structuring, indicating the 
possible existence of network motifs.

@@

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/most-abundant-topologies}
 \caption{\bf Most frequent network topologies.}
 In this figure we show the most frequent network topologies, which contain the 
 I3-FFL network motif. Top left of each box is presented the network topology 
 number in order of abundance. Top center is presented the number of GRNs 
 belonging to that network topology. Circled in red are the input nodes of 
 I3-FFL in network topologies 1 to 5, that display the striped pattern of gene 
 expression using three nodes, in contrast, network topology number 9 displays 
 the same phenotype using a two-node GRN.
 \label{fig:topol}
\end{figure}

The neutral network shows that most of the network topologies are grouped into a
connected graph. This graph is composed of 639 nodes that are accessible by one 
mutational step. The 15 most abundant network topologies, with the exception of 
number 10, are located in this graph (Fig~\ref{fig:neutral-network}). We also 
found nodes inside the neutral network that had more connections between them 
than with other nodes, generating thus clusters of nodes, interestingly in 
almost every cluster is present one or more of the most abundant network 
topologies. In addition, we found that the first eight most abundant network 
topologies were located in the largest cluster.

@@

The degree distribution of this network seems to follow a Poisson distribution
(Pearson’s $\chi^2 = 12.8902$, $\text{p-value} = 0.115684$) characteristic of a 
Erdös-Rényi network \cite{Erdos1959}. This kind of network presents a great 
number of nodes with low degree and a few nodes with high degree 
(Fig~\ref{fig:deg-dist}). Nodes corresponding to topologies number one and three
showed the highest degree ($deg(1) = deg (3) = 14$). Curiously, nodes with 
higher degree resulted to be also those network topologies with higher abundance
(Fig~\ref{fig:ab-deg}), indicating a high robustness to topological mutations as
well as robustness to mutations in the parameter values. This relationship 
between abundance and node degree was corroborated by the Spearman correlation 
($r_S = −0.353752$, $p = 6.2933\times10^{-23} $).

@@

Hierarchical clustering of spatiotemporal expression profiles showed that GRNs
with topologies that seemed to be very different have indeed very similar
expression dynamics. We expected to find GRNs with the same topology to group
together but instead we found GRNs with different topologies grouped as sister
taxa (Fig~\ref{fig:exp-tree}). More interestingly, some GRNs displaying very
abundant network topologies are grouped together with GRNs displaying very
infrequent network topologies.

@@

Although the topology 9 consists of only two genes, its mechanism of expression
is dynamically more complex than that of other topologies such as the simple
I3-FFL. Initially the expression pattern of the gene A is uniform in the
morphogenetic field, but later it is shaped by the morphogen gradient in such a
way that two important events occur, in first place the overall expression
level of the gene A is decreased by the inhibition of the morphogen, releasing
gene C from its repression by gene A; in second place, appears a non-homogeneous
spatial distribution in which gene A is expressed at higher levels in the cells
at the rightmost part of the morphogenetic field.\\

These two events conduce to the repression of the gene C by the gene A in the
rightmost part of the morphogenetic field and by the morphogen in the leftmost
part of the morphogenetic field, thus generating a lightly higher level of
expression of gene C in the center of the field. Because gene C activates its
own expression, a band of expression begins to form. This band of expression
then begins to activate the gene A also in the center of the field
(\nameref{S2_Fig}).
% (Fig~\ref{fig:topol9-dyn}).

@@

In this way two bands of expression are formed that depend on each other. The
high level of gene C in the middle of the field counteract its inhibition by
gene A, while in the ends of the bands the inhibition of gene A on gene C
overcomes the self-activation of gene C, preventing the expression of the
latter to extend out of the middle of the field. In turn, the band of
expression of gene A can only exist in the middle of the field because its
expression depends on gene C.

@@

We found the presence of network motifs in the GRNs produced by the genetic
algorithm. However, we also found a great diversity of network topologies that
could produce a striped gene expression pattern.\\

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Gene regulatory networks producing a striped band of gene 
expression} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Juan Camilo Arboleda-Rivera\textsuperscript{1\Yinyang},
Gloria Machado-Rodríguez\textsuperscript{1\Yinyang},
Boris Rodríguez-Rey\textsuperscript{2\textcurrency},
Jayson Gutiérrez\textsuperscript{3},
% Name5 Surname\textsuperscript{2\ddag},
% Name6 Surname\textsuperscript{2\ddag},
% Name7 Surname\textsuperscript{1,2,3*},
% with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Instituto de Biología, Universidad de Antioquia, Medellín,
Antioquia, Colombia
\\
\textbf{2} Instituto de Física, Universidad de Antioquia, Medellín, Antioquia,
Colombia
\\
\textbf{3} Flanders Marine Institute (VLIZ), Wandelaarkaai 7, 8400 Oostende,
Belgium
% \textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}

\subsection*{Background}
One of the central problems in developmental and synthetic biology is to 
understand how cells process external cues on its relative position to other 
cells in a developing embryo, a tissue or in a petri dish and transform such
information into a coherent response, e.g. terminal differentiation state. It 
has been stated that a gradient of concentration of certain molecule is enough 
to provide such positional information and produce cell differentiation by 
interacting with the cell's gene regulatory networks (GRNs).

\subsection*{Methods}
In this study we used a genetic algorithm in order to explore the genotypic 
space of GRNs capable of producing a band expression pattern in a group of 
cells.

\subsection*{Results}
Out of 2061 GRNs selected using the genetic algorithm, 714 groups of 
topologies were found that were able to interpret a morphogen input. The 
main network motif involved in the formation of the band expression pattern 
was the type 3 Incoherent Feed Forward Loop (I3-FFL); we also present here 
other network topologies that have not been previously reported.

\subsection*{Conclusions}

@@

Our understanding of how GRNs process information has increased thanks to the
concept and theory of network motifs, which are defined as patterns of
interconnections occurring in networks at numbers that are significantly higher
than those in randomized networks \cite{Milo2002}. This novel approach, along
with mathematical modeling and computational methods have been shown to be a
powerful tool to study GRNs and the predictions of studies using these methods
have been validated by experimental studies
\cite{Kalir2005,Kalir2004,Mangan2003a,ODonnell2005}, and some of them 
have used synthetic GRNs to produce artificial cell differentiation in bacteria
cell cultures~\cite{Basu2005}.

@@

The classification of GRNs showed that our initial set of 2061 GRNs can be 
grouped in a number of 714 distinct network topologies, each one of them
containing a different number of GRNs (Fig~\ref{fig:distopol}). This abundances
distribution was not produced by chance (Shannon entropy = 8.52963, 95\%
confidence interval = [10.8915, 10.9275]), and instead it displays some level 
of structure, indicating the existence of network motifs.

@@

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/most-abu-topol}
 \caption{\bf Most frequent network topologies.}
 Most frequent network topologies contain the I3-FFL network motif. Topologies 
 are numbered in order of decreasing abundance.
 The number of GRNs belonging to each network topology is presented below 
 topology number.
 Network topology number 9 produces the striped phenotype using a two-node GRN.
 \label{fig:topol}
\end{figure}

The neutral network (a metagraph in wich each node is a network topology) 
shows that most of the network topologies are grouped into a connected graph.
This graph is composed of 639 nodes that are accessible by one mutational step.
The 15 most abundant network topologies, with the exception of 
number 10, are part of this graph (Fig~\ref{fig:neutral-network}). We also 
found nodes inside the neutral network that had more connections between them 
than with other nodes, generating clusters of nodes, interestingly in 
almost every cluster is present one or more of the 15 most abundant network 
topologies. In addition, we found that the first eight most abundant network 
topologies were located in the largest cluster.

@@

The degree distribution of this network seems to follow a Poisson distribution
(Pearson’s $\chi^2 = 12.8902$, $\text{p-value} = 0.115684$) characteristic of a 
Erdös-Rényi network \cite{Erdos1959}. This kind of network presents a great 
number of nodes with low degree and a few nodes with high degree 
(Fig~\ref{fig:deg-dist}). Nodes corresponding to topologies number one and three
showed the highest degree ($deg(1) = deg (3) = 14$). Curiously, nodes with 
higher degree resulted to be also those network topologies with higher abundance
(Fig~\ref{fig:ab-deg}). This relationship between abundance and node degree 
was corroborated by the Spearman's rank correlation coefficient 
($r_S = −0.353752$, $p = 6.2933\times10^{-23} $).

@@

Hierarchical clustering of spatiotemporal expression profiles showed that GRNs
with topologies that seemed to be very different have indeed very similar
expression dynamics. We expected to find GRNs with the same topology to group
together but instead we found GRNs with different topologies grouped as close
neighbors (Fig~\ref{fig:exp-tree}). More interestingly, some GRNs displaying 
very abundant network topologies were grouped together with GRNs displaying 
very infrequent network topologies.

@@

Although the topology 9 consists of only two genes, its mechanism of expression
is dynamically more complex than that of other topologies such as the simple
I3-FFL. Initially the expression pattern of the gene A is uniform in the
morphogenetic field, but later it is shaped by the morphogen gradient in such a
way that two important events occur, in first place the overall expression
level of the gene A is decreased by the inhibition of the morphogen, releasing
gene C from its repression by gene A; in second place, it appears a 
non-homogeneous spatial distribution in which gene A is expressed at higher 
levels in the cells at the rightmost part of the morphogenetic field.\\

These two events conduce to the repression of the gene C by the gene A in the
rightmost part of the morphogenetic field and by the morphogen in the leftmost
part of the morphogenetic field, thus generating a slightly higher level of
expression of gene C in the center of the field. Because gene C activates its
own expression, a band of expression begins to form. This band of expression
then begins to activate the gene A also in the center of the field
(\nameref{S2_Fig}).
% (Fig~\ref{fig:topol9-dyn}).

@@

In this way two bands of expression that depend on each other are formed. The
high level of gene C in the middle of the field counteract its inhibition by
gene A, while at the ends of the bands the inhibition of gene A on gene C
overcomes the self-activation of gene C, preventing the expression of the
latter to extend out of the middle of the field. In turn, the band of
expression of gene A can only exist in the middle of the field because its
expression depends on gene C.

@@

We found the presence of network motifs in the set of GRNs produced by the
genetic algorithm. However, we also found a great diversity of network 
topologies that could produce a striped gene expression pattern.\\

That the most abundant topologies were found to be part of highly connected 
clusters in the neutral network indicates that these topologies are both robust 
to quantitative changes in the strength of interactions and diffusion and 
degradation parameters as well as to qualitative changes in the topology 
itself.
